---
title: "stat442final project"
author: "Robin"
date: "2022-12-08"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
Every NBA season, there will be a selection of the Most Valuable Player (MVP) award. It is one of the most prestigious accolades that an NBA player can receive in his career. The selection is conducted by independent media members who are not affiliated with teams nor players. Each member selects players in a weighted voting system. From 2010, public can also participate in voting the MVP. However, there are so many irrational factors such as public image and scandals that can affect the selection of a player. In my project, I will only focus on the performance of playing basketball, using 4 visualization, based on the data NBA_Players_Boxscore, to choose my MVP.


# Data preparation
There are over 300 players in the league, while most of them are not eligible for being MVP, so I will filter the data first. The eligible players should be:  
- starters for every game played   
- average playing time per game more than 25 minutes  
- scores over 23 points per game    
```{r,warning=FALSE}
nba <- read.csv("C:/Users/76506/Desktop/stat442 final/NBA_Player_Boxscore_2021-22.csv",
                stringsAsFactors=TRUE)
library(dplyr)
# average playing time per game more than 25 minutes
df <- nba %>%  group_by(athlete_display_name) %>% 
  summarise_at(vars(min),list(avgtime = mean)) %>%
  filter(avgtime > 25)
nba <- subset(nba,athlete_display_name %in% df$athlete_display_name)
# starters for every game played
df <- nba %>% group_by(athlete_display_name) %>% filter(starter == FALSE)
nba <- subset(nba,!athlete_display_name %in% df$athlete_display_name)
# scores over 20 points per game
df <- nba %>% group_by(athlete_display_name) %>%   
  summarise_at(vars(pts),list(avgpts = mean)) %>%
  filter(avgpts > 23)
nba <- subset(nba,athlete_display_name %in% df$athlete_display_name)
# qualified players
players <- unique(nba$athlete_display_name)
length(players)
```
After filtering by basic requirements, there are 19 players left for later discussion. 

# Visualization
# Viz1: Wildcard (a huxtable of players' average rebounds and assists sorted by mean points)
PPG: points per game
RPG: rebounds per game
APG: assists per game
```{r,warning=FALSE}
library(huxtable)
library(ggplot2)
library(knitr)
df <- nba %>% group_by(athlete_display_name) %>%
summarise('player name' = athlete_short_name[1],
'position' = athlete_position_name[1],
'PPG' = mean(pts),
'RPG' = mean(reb),
'APG' = mean(ast))%>%
arrange(-`PPG`) %>%
select(-`athlete_display_name`)
mytable <- df %>% as_huxtable() %>%
set_background_color(evens,everywhere,"grey")%>%
set_outer_borders()%>%
set_bottom_border(1,)%>%
set_bold(1,everywhere)
mytable
```  


We can find that players such as Giannis, Lebron, and Jokic have not only scored many points, their rebounds and assists are also better than most of others. Because of the basketball rules, players on different positions have different data structure. Generally, centers are responsible for getting rebounds, guards have more assists and forwards are more comprehensive. Therefore, we need to facet the players by their position. 

# Viz2 categorical(facet by positions)

```{r,warning=FALSE}
mygraph <- nba %>% ggplot(aes(x=reb,y=pts)) +
geom_point() +
geom_smooth(method = "loess",
se = TRUE,col = "red") +
facet_wrap(~athlete_position_abbreviation) +
labs(title = "NBA Player Points vs Rebounds by position",
x = "Rebounds",
y = "Points")
mygraph
```
As shown in the graph, because of the basketball rules, players on different positions have different data structure. Generally, centers are responsible for getting rebounds, guards have more assists and scores, forwards are more comprehensive. Therefore, it is not wise to compare players' stats directly.

# Viz3 2D (dot plots for different positions)
Let's plot three dot plots, one for centers, one for forwards, one for guards, and find the best player in each position. Let's use the first 2 or 3 players with highest PPG in each position, based on the huxtable from Viz1.
```{r,warning=FALSE}
library(ggpubr)
library("patchwork")
df_forward <- subset(nba,athlete_display_name %in% c('LeBron James', 'Giannis Antetokounmpo','Kevin Durant'))
df_center <- subset(nba,athlete_display_name %in% c('Joel Embiid', 'Nikola Jokic'))
df_guard <- subset(nba,athlete_display_name %in% c('Trae Young', 'Ja Morant','Kyrie Irving'))
# box-dot plot
plot_center <- ggplot(df_center,
                      aes(x=athlete_short_name, y=reb)) + geom_boxplot()+
  geom_dotplot(binaxis='y', stackdir='center',binwidth = 0.5)+
  labs(x = "player name",y ="rebounds")
# violin-dot plot
plot_forward <- ggplot(df_forward, aes(x=athlete_short_name, y=pts)) +geom_violin(trim = FALSE)+
  geom_dotplot(binaxis='y', stackdir='center',binwidth = 1.5)+
  labs(x = "player name",y ="points")
# notched-box plot
plot_guard <- ggplot(df_guard, aes(x=athlete_short_name, y=ast)) + geom_boxplot(notch = TRUE)+
  geom_dotplot(binaxis='y', stackdir='center',binwidth = 0.5)+
  labs(x = "player name",y ="assists")
ggarrange(plot_center, plot_forward, plot_guard,
          labels = c("center", "forward", "guard"),
          ncol = 2, nrow = 2) + 
  plot_annotation(title = "dot plots faceted by positions") & 
  theme(plot.title = element_text(hjust = 0.5))
```
